Sustainable Pathways for Irish homes in changing TIMES

### 2021-07-26 manuscript.tex
Done for the day, rereading - at Ireland residential sector

\documentclass[preprint,12pt]{elsarticle}
%\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{pgfplots}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{siunitx}
\usepackage{eurosym}
\usepackage{multirow}
\usepackage{xcolor} 
\usepackage{colortbl}
\usepackage{subcaption}
\usepackage{blindtext}
\usepackage{lineno,hyperref}
\modulolinenumbers[5]
\DeclareSIUnit{\EUR}{\text{\euro}}
 \pgfplotsset{compat=1.17}
%\usepackage[table]{xcolor}
\usepackage{biblatex}

%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}
\addbibresource{references.bib}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.

\journal{Energy and Buildings}

@@

\begin{abstract}
%% Text of abstract
%Ireland's residential sector has particularly underachieved in previous climate targets. Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector. The disadvantageous starting position has impeded residential decarbonisation progress thus far. \par
%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy. \par
%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
\par The Irish residential sector has proved hard to decarbonise, yet recently Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018. There is a need for a new state of the art model to provide insights on the optimal decarbonisation pathway for the Irish energy system. This paper reviews residential and energy system models and describes the methodology applied to the development of the residential sector in the new TIMES Ireland Model (TIM). The paper includes some results to provide a discussion on some of the insights and highlights the insights which may need to be further investigated in the residential sector.

@@

Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways and interpreting the results. Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector. While also examining how researchers have previously used bottom-up residential models and Energy System Optimization Models (ESOMs) to analyse the residential sector.
Section \ref{LitR} describes the methodology used to develop the residential sector. The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted. 
Section \ref{Results} outlines the six sceanrios used and displays the energy system and residential results. Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models. Lastly, section \ref{Conclusion} draws some conclusions on the model, results and discussion on future work.

@@

ESOMs have been used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-BasedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore certain economic sector in higher granularity.

TIMES has been extensively used in both the UK and Ireland, and there are several examples using TIMES to focus on the residential sector. TIMES assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. With this in mind, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices, this identifies perfect foresight as a weakness in TIMES. While Li's research has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.
 \par Many sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation. Using the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.}
Leibowicz et al.\cite{Leibowicz2018OptimalServices} explores the optimal decarbonisation of the residential sector by examining electricity generation and storage strategies. The results find net present costs in the \textit{``Policy scenarios"} ( $CO_2$ declines linearly by 90\% in 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ). However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock, highlighting the importance of energy building regulation as thermally efficient building stock can reduce the decarbonisation cost. 
District Heating is a proven low carbon heating technology, however it is almost non-existant and has not been included in previous ESOM in Ireland. Denmark are leading experts in this area and Denmark's district heating system has been comprehensively examined. One study by 
Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} used the Balmorel model to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark. Exogenous user constraints applied to the model, based on the current Danish policy. The results find that air-to-water heat pumps are installed in non-district heating areas, while the system benefits of adding individual heat storage to the heat pumps are only moderate, the main benefit is a reduced need for peak/reserve capacity investments. This identifies individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments in areas where district heating is not available.
Another technology which is proven but under utilised in Ireland, is photovoltaic (PV) solar panels. While Ireland's policy and building regulations have encouraged the uptake of PVs, however the market share of PV systems remains low, and as the cost continues to fall they can alter optimal decarbonisation pathways.   
Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.

@@

%The Cambridge Housing Model (BEIS) ses English Housing Survey data, coupled to a SAP ( UK's BER equivalent ) energy calculator, to estimate energy use and CO2 emissions for all homes in England, broken down by final use. The CHM reads in the EHS dwelling for each case and performs building physics calculations to determine energy consumption and associated CO2 emissions, by use and by fuel type. Multiplying the energy use and CO2 emissions by the associated weighting and summing across all cases gives total values for England. Housing Data is cleaned, Climate data for region is 8 energy services - space heating ( main \& secondary ), water heating, space cooling, lighting, electrical appliances, cooking and pumps \& fan. 6 fuels - gas, oil, solid, biomass, electriicty and renewable

@@

ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.

In 2014, Rogan et al. \cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM, to the Long range Energy Alternatives Planning system (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland. The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS. Three scenarios are exploorange, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario. The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020. In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compared to 2008. The share of space and water heating decreased from 82 \% to 77 \% in that period. In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario. The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits. This results in an energy consumption reduction of 16.5 \% in 2020, compared to the reference scenario.Rogan et al. discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector. The importance for the policy to not only look at the number of retrofits, but also the depth of retrofit work, is shown by the difference in scenarios. The model also found considerable scope for energy savings to be made through improved building regulations.

In 2018, Thellufsen et al. \cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland. The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN. The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland. A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal. The final results find that district heating compared fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal. Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.

\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models, rather than energy system models specifically examining the residential sector. Nevertheless, the residential models have provided insights on the input data and methodology approach used in Ireland, some of which are outlined below.

Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector. This was the first comprehensive analysis of the Irish residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}. In 2000, Clinch and Healy published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating. A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts. Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component into the computer model. The modelling process involved developing a computer model of the energy performance of the existing dwelling stock. It was noted that the energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions. 
The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.

In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on the energy efficiency of space and water heating by examining technology and building fabric efficiency. The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations. Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.  
\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock. A base geometry according to age bands and thermal characteristics was created in the study. The heat load for a statistical occupancy and year was established. The default values used in some of the BER database variables where highlighted. For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey found 18\% of detached housing in Ireland had no roof insulation. Ahern et al. found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area. Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset. The model thermal improvement measures resulted in a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback. Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.

\par In 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock. The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for BER calculation and checking building regulation compliance. The DEAP model used 175 archetype dwellings ( 5 building types, 7 energy bands, and 5 wall construction types ) and 6 different retrofit measures were applied. BER database filters were used in this model, the BER filters from this study were updated and applied in TIM. A sensitivity analysis is performed and it was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift. Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.

The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}. ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs. Mac Uidhir et al. specifically examined the Better Energy Homes (BEH) scheme. The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ). The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario. This shows the the type of retrofit and type of house can have a large affect on the energy saved. 
Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is negligible.

Finally, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system. Some of the main models used in the NEMF include:
\begin{itemize}
\item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{2021PLEXOSExemplar}
\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingOpportunity}
\begin{itemize}
\item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
\end{itemize}
\item{BioHEAT is a techno-economic simulation model of bioenergy and heat, used by SEAI to support policy decision in Ireland.}\cite{Durusut2018BioHEAT:Sectors}
\end{itemize}

@@

The models described in this section include a range of national ESOMs and bottom-up residential models. The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating the health effects of retrofitting, and integration of bioenergy and heat. The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the growing BER database is providing high data granularity input to these models. The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures. 
The bottom-up residential models are typically simulation models looking at a specific policy. The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL. 
There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector. The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport. The use of ESOM to explore DH in Ireland allows for insights which bottom-up simulation models cannot provide. The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.

\subsection{Ireland: Residential Sector}
\label{Intro:IreRSD}

>>>>

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
%\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{pgfplots}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{siunitx}
\usepackage{eurosym}
\usepackage{multirow}
\usepackage{xcolor} 
\usepackage{colortbl}
\usepackage{subcaption}
\usepackage{blindtext}

\DeclareSIUnit{\EUR}{\text{\euro}}
 \pgfplotsset{compat=1.17}
%\usepackage[table]{xcolor}
%\usepackage[backend=bibtex]{biblatex}

%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}
%\bibliographystyle{elsarticle-num}
%\usepackage{cleveref}
%\usepackage{hyperref}
%\usepackage[style=numeric, sorting=none]{biblatex}  %backend=biber is 'better'
%
\addbibresource{TIM_V11.bib}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.

\journal{Energy and Buildings}
%\bibliographystyle{elsarticle-num}

@@

\begin{abstract}
%% Text of abstract
%Ireland's residential sector has particularly underachieved in previous climate targets. Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector. The disadvantageous starting position has impeded residential decarbonisation progress thus far. \par
%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy. \par
%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
\par The Irish residential sector has proved hard to decarbonise, yet recently Ireland has set one of the most ambitious climate targets into law, a 51 \% greenhouse gas reduction in 2030, compared to 2018. There is a need for a new up to date and state of the art model to provide insights on the optimal decarbonisation pathway for the Irish energy system. This paper describes the methodology applied to the development of the residential sector in the new TIMES Ireland Model (TIM). Results are analysed to provide input to the discuss insights from the residential sector in TIM.

@@

The layout of this paper is as follows: 
\begin{itemize}
    \item Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy, and distinctive residential sector. While also examining how researchers have previously used bottom-up residential models and Energy System Optimization Models (ESOMs) to analyse the residential sector.
    \item Section \ref{LitR} is the core section of this paper and explains the methodology used to develop the residential sector. The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted. 
    \item Section \ref{Results} outlines the scenarios used and displays results from the residential sector.
    \item Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models.
    \item Lastly, section \ref{Conclusion} draws some conclusions and discusses possible future work.
\end{itemize}

@@

 ESOMs have been used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOMs currently used include MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOMs cover all energy-related sectors, they can also be used to specifically explore certain economic sectors in higher granularity. This section gives some examples of residential sector studies using ESOMs.

TIMES has been extensively used in both the UK and Ireland, and there are several examples using TIMES to focus on the residential sector. TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. With this in mind, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies. A nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. Despite the perfect foresight assumption being essential to finding a solution, it has the potential to be a weakness by widening the gap between the model and reality. 
\par A common theme across many ESOMs shows different sectors are increasingly reliant on electricity generation as lower carbon intensity electricity is considered a cornerstone in optimal decarbonisation. Using the Open Source Energy Modeling System (OSeMOSYS), Leibowicz et al. \cite{Leibowicz2018OptimalServices} explores the optimal decarbonisation of the residential sector by examining electricity generation and storage strategies. The results find net present costs in the \textit{``Policy scenarios"} ( $CO_2$ declines linearly by 90\% in 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ). However, the percentage increase in net present cost diminishes with a more thermally efficient building stock, highlighting the importance of energy building regulation as thermally efficient building stock can reduce residential decarbonisation costs. 
\par District Heating is a proven low carbon heating technology, however it is rarely researched in Ireland and has not been included in previous ESOM in Ireland. However, Denmark's district heating system has been comprehensively examined. One study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} used the Balmorel model to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark. The results find that air-to-water heat pumps are installed in non-district heating areas, while the system benefits of adding individual heat storage to the heat pumps are only moderate, the main benefit is a reduced need for peak/reserve capacity investments. This identifies individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments in areas where district heating is not available.
\par While Ireland's building regulations have encouraged the uptake of photovoltaics (PVs), the market share of PV solar panel systems remains low. The low amount of direct sunlight means Ireland's climate is not ideal for PVs but as the cost of PVs continues to fall, the potential role of PVs in  residential decarbonisation rises. 
In one study, Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model to explore the feasibility of PV systems installed at the household level in combination with batteries in Samsø.

@@

@@

ESOMs have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2009DevelopingUniversity.}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are outlined below.

In 2014, Rogan et al. \cite{Rogan2014LEAPsSystem} linked OSeMOYOS to the Long Range Energy Alternatives Planning system (LEAP) simulation model to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland. Three scenarios were explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario. Where the reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020 and the results show that residential total energy consumption rose 6.7 \% in 2020, compared to 2008. In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario. The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits. This results in an energy consumption reduction of 16.5 \% in 2020, compared to the reference scenario. Rogan et al. discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector. The importance for the policy to not only look at the number of retrofits, but also the depth of retrofit work, is shown by the difference in scenarios. The model also found considerable scope for energy savings to be made through improved building regulations.

In 2018, Thellufsen et al. \cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland. The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN. The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland. A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates. In all cases, the inclusion of district heating was found to be favourable. The results find the conversion of 37 \% of the Irish heat demand to district heating to be optimal. Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.

\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models, rather than energy system models specifically examining the residential sector. Nevertheless, the residential models have provided insights on the input data, methodology, and results, and some studies are outlined below.

Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector. This was the first comprehensive analysis of the Irish residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}. In 2000, Clinch and Healy published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating. A cost-benefit model (CBM) was fitted to the EAM and the results under the predicted scenario, showed energy savings outweighed programme costs, representing a benefit-cost ratio of 1.7. The benefits of reduced sickness, hospitalisation, drugs, and mortality were later quantified and this moved the benefit-cost ratio up to 3.0. The same model was later used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} to represent the human behavioural component in the computer model.

\par
In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on the energy efficiency of space and water heating by examining technology and building fabric efficiency. The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations. The paper acknowledges that retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperature effect on energy efficiency improvements.  
\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures of the existing detached, oil centrally heated, rural housing stock. A base geometry according to age bands and thermal characteristics was created in the study. The default values used in some of the BER database variables where highlighted. For example, there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey finding 18\% of detached housing in Ireland had no roof insulation. Ahern et al. found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area. Concluding that the majority of thermal energy efficiency progress in Irish dwellings is being offset. The model results show that thermal improvement measures resulted in a reduction of running costs and $CO_2$ emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback. Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures. This study highlights the low hanging fruit of pre-1980 oil centrally heated, rural housing stock.

\par In 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock. The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for BER calculation and checking building regulation compliance. The DEAP model used 175 archetype dwellings ( 5 building types, 7 energy bands, and 5 wall construction types ) and 6 different retrofit measures were applied. The BER filters from this study were updated and applied in TIM. A sensitivity analysis performed found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift. Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.

\par The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}. ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption using EN 13790 to simulate the potential for improved energy efficiency gains within the existing retrofit programs. Mac Uidhir et al. specifically examined the Better Energy Homes (BEH) scheme. The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ). The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario. This shows the the type of retrofit and type of house can have a large affect on the energy saved.

\par Finally, it is worth mentioning Sustainable Energy Authority of Ireland's (SEAI’s) National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system. The SEAI is Ireland's national energy authority and they play a key role in assisting consumers to reduce energy. Some of the main models used in the NEMF include:
\begin{itemize}
\item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning} \cite{2021PLEXOSExemplar}
\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingOpportunity}
\begin{enumerate}[i.]
\item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
\end{enumerate}
\item{BioHEAT is a techno-economic simulation model of bioenergy and heat} \cite{Durusut2018BioHEAT:Sectors}
\end{itemize}

@@

The models described in this section include a range of national ESOMs and bottom-up residential models. The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating the health effects of retrofitting, and integration of bioenergy and heat. The thermal efficiency of dwellings is commonly researched in Irish models, due to the poor thermal efficiency of the dwelling stock and the growing BER database is providing high data granularity input to these models. The internal temperature is becoming more frequently analysed in the studies, a recurring theme is the wide range of internal temperatures. 
The bottom-up residential models are typically simulation models looking at a specific policy. The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL, with much higher computational power. 
There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector. The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport. The use of ESOM to explore DH in Ireland allows for insights which bottom-up simulation models cannot provide. The connection between waste heat in electricity and industry can provide better insights on the optimal solution for DH in commercial and residential sectors.

\subsection{Ireland: Residential Sector}
\label{Intro:IreRSD}
%PICKUPHERE
Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns in Ireland's residential sector. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{Eurostat2021DistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{Eurostat2021ShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of $CO_2$ emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousing2021Gov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020EnergySEAI}.
